---
title: "Association Study with lazyGas"
author:
  - name: "Tomoyuki Furuta"
    affiliation: Institute of Plant Science and Resources, Okayama University, Okayama, Japan
    email: f.tomoyuki@okayama-u.ac.jp
date: "July 9, 2026"
package: lazyGas
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Genetic Association Study with lazyGas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
This vignette demonstrates the process of conducting a genetic association study using the `lazyGas` package. We will go through the steps of loading input files, preprocessing data, performing association analysis, and visualizing results.

Companion storage for scan, peak, recalc, and candidate outputs is written by default to a Parquet-backed directory next to the GDS file (or optionally SQLite or the legacy GDS `lazygas/` subtree); see the [Where association results are stored](#where-association-results-are-stored) section. Optional refinement of candidate genes with `searchCandidateGenes()` uses phrase-aware keyword matching (stopwords ignored; word-boundary / phrase patterns).

Since v0.5.0, lazyGas also provides a **variant viewer** (`getVariantViewerData()`, `plotVariantViewer()`) and an **interactive dashboard** (`makeInteractiveDashboard()`) that links candidate genes to per-gene variant plots in a single HTML file. Runnable demos are available via `runDashboardDemo()` and, for v0.6+ features, `runMvpDemo()` using bundled files in `inst/extdata/`. Since v0.7.0, the **phenotype explorer** (`runLazyGasExplorer()`, `runExplorerDemo()`) ranks candidates by multi-source evidence with optional local LLM support. Since v0.8.0, GWAS **`Coef.*` columns are stored in original phenotype units** for continuous traits (see [Interpreting scan coefficients](#interpreting-scan-coefficients-coef)). Since v0.9.0, the Shiny explorer also embeds **GWAS overview** and **Locus / variants** plots (same helpers as the HTML dashboard). Since v0.9.1, a separate **pipeline runner** (`runLazyGasRunner()`) runs `buildLazyGas()` → `assignPheno()` → `runLazyGas()` from filesystem paths before handing off to the explorer. Since v0.9.2, explorer sidebar controls are tab-scoped with a dedicated Local LLM diagnostics tab and busy-state progress/locking for rank/chat actions.

Load the lazyGas package.
```{r message=FALSE, results="hide", warning=FALSE}
library(lazyGas)
have_demo_ann <- FALSE
```
<br><br>

## Setting up your environment

Use this checklist before running the examples below or the bundled demos.

### System and core dependencies

| Requirement | Notes |
|-------------|-------|
| R ≥ 4.1 | Recommended; cache paths use `tools::R_user_dir()` |
| GNU make, C++11 | Required to build some dependencies (`SystemRequirements` in `DESCRIPTION`) |
| **GBScleanR** | Hard dependency — install from [GitHub](https://github.com/tomoyukif/GBScleanR) before `lazyGas` |
| **arrow** | Parquet companion storage (installed with lazyGas imports) |

Install CRAN dependencies and lazyGas:

```{r eval=FALSE}
install.packages(c(
  "ggplot2", "arrow", "jsonlite", "dplyr", "cowplot",
  "parameters", "gaston", "vcfR", "plotly", "htmltools",
  "reactable", "devtools"
))
# Install GBScleanR from GitHub, then:
devtools::install_github("tomoyukif/lazyGas", build_vignettes = TRUE)
```

From a local source tree during development:

```{r eval=FALSE}
devtools::load_all("/path/to/lazyGas")
```

### Optional packages by feature

| Package | Feature |
|---------|---------|
| `shiny` | `runLazyGasRunner()` pipeline UI; `runLazyGasExplorer()` (GWAS / locus / ranking / chat) |
| `base64enc` | Static Manhattan PNG in HTML / explorer |
| `DBI`, `RSQLite` | `lazygas_store = "sqlite"` |
| `processx` | Background Ollama from R |
| `httr2`, `curl` | Local LLM HTTP API |
| `rentrez` | PubMed literature evidence (explorer; optional) |

`plotly`, `htmltools`, and `reactable` are package **Imports** (HTML reports and
explorer plots).

```{r eval=FALSE}
install.packages(c("shiny", "base64enc", "DBI", "RSQLite", "processx", "httr2", "curl"))
```

### Optional local LLM (phenotype explorer)

Evidence ranking works without an LLM. For richer query parsing and explanations:

1. **Apptainer / Singularity** (recommended on clusters; no host `ollama` binary):
   - Install Apptainer or Singularity; ensure `apptainer` or `singularity` is on `PATH`.
   - One-time in R: `configureLazyGasOllama(model = "llama3.2:3b")` then `startLocalLLM(model = "llama3.2:3b")`.
   - Shell: `bash inst/apptainer/build_ollama_sif.sh` (see [Phenotype explorer and local LLM](#phenotype-explorer-and-local-llm-apptainer-official-setup)).
2. **Host Ollama**: install from https://ollama.com and run `startLocalLLM()`.

Cache and models default to `lazyGasOllamaHome()` (override with `LAZYGAS_OLLAMA_HOME`).

### Verify installation

```{r eval=FALSE}
out_dir <- tempfile("lazygas_check_")
runDashboardDemo(out_dir = out_dir)
file.exists(file.path(out_dir, "lazygas_dashboard.html"))
```

<br><br>

## Loading Input GDS file and Building a LazyGas Object
The lazyGas package was primarily designed to take as input a GDS file created 
using the `r Biocpkg("GBScleanR")` package. If you have a VCF file for 
regression analysis, you can convert it to a GDS file using the gbsrVCF2GDS()
function in GBScleanR.  
Load a sample GDS file and build a `LazyGas` class object.
```{r message=FALSE, results="hide", warning=FALSE}
gds_fn <- system.file("extdata", "sample.gds", package = "lazyGas")
lg <- buildLazyGas(gds_fn = gds_fn, load_filter = TRUE, overwrite = FALSE)
```
If you applied filtering on markers and/or samples using the functions in the 
`r Biocpkg("GBScleanR")` package, please set `load_filter = TRUE` that apply the
same filters on the given data during building a `LazyGas` class object.

The setting `overwrite = FALSE` ensures that existing lazyGas outputs in the chosen
result store are preserved. If `overwrite = TRUE`, those outputs will be removed.
This setting has no effect when creating a `LazyGas` class object for the first
time, but it does apply when reloading the GDS file for reanalysis.  
<br>

### Where association results are stored
Genotypes and marker coordinates stay in the GDS file. Since lazyGas v0.5.0,
association outputs (`scan`, `peakcall`, `recalc`, `candidate`, and related
metadata) are written to a **companion store** by default, not into a `lazygas/`
subtree inside the GDS. Choose the backend with `lazygas_store` in
`buildLazyGas()`:

| `lazygas_store` | Location |
|-----------------|----------|
| `"parquet"` (default) | `{gds_stem}.lazygas/` (Parquet files and `meta.json`) |
| `"sqlite"` | `{gds_stem}.lazygas.sqlite` (requires **DBI** and **RSQLite**) |
| `"gds"` | `lazygas/` subtree inside the GDS file (legacy layout) |
| `"auto"` | GDS subtree if it already exists; otherwise Parquet companion |

For the sample data above, scan results are written under
`sample.lazygas/scan/` next to `sample.gds` when the default is used.
<br>
  
If you have genotype information that are not stored in a GDS file, you can build 
a `LazyGas` class object by supplying a named list including the necessary 
information to the `create_gds` argument of the `buildLazyGas()` function.
```{r message=FALSE, results="hide", warning=FALSE}
# Set a temporary file path for a GDS file.
temp_dir <- tempdir()
sample_gds <- tempfile("sample", temp_dir, ".gds")

# Now we assume genotype and marker data were provided as a matrix or 
# data.frame that have been loaded from CSV files, although here we use data 
# obtained from the GDS file for ease of handling as example usage.
genotype <- getGenotype(object = lg, node = "raw")
dosage <- getGenotype(object = lg, node = "dosage")
haplotype <- getHaplotype(object = lg)
snp.chromosome <- getChromosome(object = lg)
snp.position <- getPosition(object = lg)
snp.allele <- getAllele(object = lg)
sample.id <- getSamID(object = lg)
snp.id <- getMarID(object = lg)

# The following values can be NULL: snp.id, snp.rs.id, and snp.allele
# Either of the following values should be specified at least: genotype, 
# haplotype, or dosage.
create_gds <-  list(genotype = genotype,
                    sample.id = sample.id,
                    snp.id = snp.id,
                    snp.rs.id = NULL,
                    snp.chromosome = paste0("chr", sprintf("%02d", as.numeric(snp.chromosome))),
                    snp.position = snp.position,
                    snp.allele = snp.allele,
                    haplotype = haplotype,
                    dosage = dosage)
lg <- buildLazyGas(gds_fn = sample_gds, 
                   create_gds = create_gds)
```
<br>

## Loading and Preprocessing Phenotype Data
Load the phenotype data.
```{r}
pheno_fn <- system.file("extdata", "pheno.csv", package = "lazyGas")
pheno <- read.csv(file = pheno_fn)
```
<br>

You need to assign phenotype data to the `LazyGas` object.
```{r}
lg <- assignPheno(object = lg,
                  pheno = pheno,
                  rename = "Fruit weight")
```
You can rename the phenotypes if you needed. The `rename` argument should be a 
character vector of the same length as the number of phenotypes in the input 
data. This renaming is useful when phenotype names in the input `pheno` 
data.frame has abbreviated or ambiguous labeling of phenotype names in the header 
of the data.frame. The phenotype names provided via the `rename` argument or 
retrieved from the data.frame header will be used in plots. Renaming is only for 
the visualization purpose and does not affect the analysis.  
<br><br>

## Visualizing Phenotype Data
Generate and save plots for phenotype data.
```{r}
# getPheno() returns a list of phenotype data. 
pheno <- getPheno(object = lg)
for(i in seq_along(pheno$pheno_names)){
  p <- suppressMessages(plotPheno(object = lg, 
                                  pheno = i,
                                  xlab = pheno$pheno_names[i]))
  print(p)
}
```
<br><br>

## Conducting Association Analysis

### Prepare a function to make a model matrix.
Association analysis can be performed using `scanAssoc()` function.
Since association analysis is a type of regression analysis that requires a 
model formula, you need to specify one accordingly.
The `scanAssoc()` function also requires a conversion function to transform the
genotype data into a model matrix that fits the model formula.
The `getGenoPerMarker()` function can be used to check the structure of the 
genotype data. If you are going to use haplotype or dosage data instead of 
genotype data, you can specify the `geno_format` argument as `haplotype` or 
`dosage`.
```{r}
g <- getGenoPerMarker(object = lg, geno_format = "genotype")
print(g)
```
<br>


```{r}
g <- getGenoPerMarker(object = lg, geno_format = "haplotype")
print(g)
```
<br>


```{r}
g <- getGenoPerMarker(object = lg, geno_format = "dosage")
print(g)
```

As shown above, genotype and haplotype data for each marker is a matrix with two
rows if you have diploid samples, whereas dosage data is a vector.
You need to prepare a function to convert those data into a model matrix for the 
regression analysis.
<br>


When you use the dosage data as input for the regression analysis,
the conversion function can be defined manually as shown below. 
```{r}
# Conversion function for dosage data assuming additive and dominant effects
conv_fun <- function(g){
  add <- g
  dom <- as.numeric(g == 1)
  out <- data.frame(add = add, dom = dom)
  return(out)
}

# For the model matrix above, the formula can be the following.
formula <- "add + dom"
```
<br>
  

Since genotype data indicates reference and alternative alleles as 0 and 1, 
you can sum the genotype values to obtain the number of alternative alleles.
If the reference allele consistently represents the reference haplotype — 
for example, when 0 and 1 in the genotype data always represent alleles 
inherited from Parent 1 and Parent 2, respectively — the number of alternative
alleles can be treated as the dosage of alternative haplotypes.
```{r}
conv_fun <- function(g){
  add <- colSums(g)
  dom <- as.numeric(g == 1)
  out <- data.frame(add = add, dom = dom)
  return(out)
}

# For the model matrix above, the formula can be the following.
formula <- "add + dom"
```

<br>

For the haplotype data, you can incorporate interaction effects of haplotypes as
follows.
```{r}
# Conversion function for haplotype data assuming additive and interaction effects
conv_fun <- function(g){
  hap1 <- colSums(g == 1)
  hap2 <- colSums(g == 2)
  hap1_hap2 <- as.numeric(hap1 == 1 & hap2 == 1)
  out <- data.frame(hap2 = hap2, hap1_hap2 = hap1_hap2)
  return(out)
}
formula <- "hap2 + hap1_hap2"

```

<br>

Although you can manually define a function for the data conversion, 
the makeConvFun() function helps you create a data conversion function. 
```{r}
# The n_levels argument specifies the number of dosage levels.
# For diploid samples, n_levels should be set to 3: 0, 1, and 2 represent 
# null-, single-, and duplex dosages, respectively.
conv_fun <- makeConvFun(geno_format = "dosage", n_levels = 3)
formula <- "add + dom"
```
  
<br>

You can check the conversion function definition
```{r}
conv_fun
```

<br>

The current implementation of the `makeConvFun()` function does not support 
to assume interaction effects of haplotypes. If necessary, please define a 
conversion function manually as shown in the previous code chunk. When you have
four haplotypes, the function definition can be the following.
```{r eval=FALSE}
# Do not run.
conv_fun <- function(g){
  hap1 <- colSums(g == 1)
  hap2 <- colSums(g == 2)
  hap3 <- colSums(g == 3)
  hap4 <- colSums(g == 4)
  hap1_hap2 <- as.numeric(hap1 == 1 & hap2 == 1)
  hap1_hap3 <- as.numeric(hap1 == 1 & hap3 == 1)
  hap1_hap4 <- as.numeric(hap1 == 1 & hap4 == 1)
  hap2_hap3 <- as.numeric(hap2 == 1 & hap3 == 1)
  hap2_hap4 <- as.numeric(hap2 == 1 & hap4 == 1)
  hap3_hap4 <- as.numeric(hap3 == 1 & hap4 == 1)
  out <- data.frame(hap2 = hap2, hap3 = hap3, hap4 = hap4, 
                    hap1_hap2 = hap1_hap2, hap1_hap3 = hap1_hap3,
                    hap1_hap4 = hap1_hap4, hap2_hap3 = hap2_hap3,
                    hap2_hap4 = hap2_hap4, hap3_hap4 = hap3_hap4)
  return(out)
}
formula <- paste0("hap2 + hap3 + hap4 + hap1_hap2 + hap1_hap3",
                  " + hap1_hap4 + hap2_hap3 + hap2_hap4 + hap3_hap4")
```
  
<br><br>

### Association Analysis
The `scanAssoc()` function performs regression analysis using a model formula
specified via the `formula` argument and a model matrix generated by a conversion 
function provided through the `conv_fun` argument, which is applied to the data 
specified by the `geno_format` argument. 
In the following case, dosage data is used when `geno_format = "dosage"` is specified.
```{r}
scanAssoc(object = lg,
          formula = formula,
          conv_fun = conv_fun,
          geno_format = "dosage")
```
  
<br>

### Interpreting scan coefficients (`Coef.*`)

For **continuous** phenotypes, `scanAssoc()` standardizes trait values internally
before regression (zero mean, unit variance). Stored **`Coef.*` columns** (for
example `Coef.add`, `Coef.dom`) are **rescaled to the original phenotype units**:
each coefficient is the expected change in the input trait per one unit of the
corresponding genotype term (e.g. per dosage allele). **P-values, FDR, and
`-log10P` are unchanged** by this rescaling.

For **binary** phenotypes, standardization is skipped; coefficients remain on
the GLM scale (log-odds).

Re-run `scanAssoc()` if you need rescaled coefficients for results saved with
lazyGas versions before v0.8.0.

```{r}
scan <- lazyData(object = lg, dataset = "scan", pheno = "Fruit weight")
if ("Coef.add" %in% names(scan)) {
  knitr::kable(scan[which.max(scan$negLog10P), c("variant_ID", "negLog10P", "Coef.add", "Coef.dom")])
}
```

<br>

For GWAS, set `method = "mlm"`. When `method = "mlm"` is specified, only 
`geno_format = "genotype"` or `geno_format = "corrected"` is accepted. 
If `geno_format = "corrected"` is specified, corrected genotype data generated by
`r Biocpkg("GBScleanR")` will be used for regression analysis. 
```{r eval = FALSE}
scanAssoc(object = lg,
          geno_format = "genotype",
          method = "mlm")
```

<br>

If you conducted association tests using another tool and have p values,
you can register those results with **`assignPvalues()`**. This writes a
**new** scan table for the phenotype; it does not update an existing
`scanAssoc()` result in place. Also pass **`coef`** (effect sizes, e.g.
`Coef.add`) and/or **`any_data`** (any other per-marker columns) when you
have them—downstream plots such as **`haploPlot()`** use stored coefficients
when present. Specify `geno_format`, `conv_fun`, and `formula` for later
pipeline steps. Repeat the call for each phenotype when you have several
traits.
```{r}
# Example: external p-values and additive coefficients (synthetic demo values)
n_ext <- nmar(lg)
ext_p <- runif(n_ext, min = 1e-6, max = 0.5)
ext_coef <- rnorm(n_ext, sd = 0.5)
lg <- assignPvalues(
  object = lg,
  pheno_name = "Fruit weight",
  p_values = ext_p,
  coef = ext_coef,
  geno_format = "dosage",
  conv_fun = conv_fun,
  formula = formula
)
```

<br><br>

## Visualizing Results
Generate Manhattan plots for association results.
```{r}
pheno <- getPheno(object = lg)
for(i in seq_along(pheno$pheno_names)){
  p <- plotManhattan(object = lg, pheno = i)
  print(p)
}

# lazyData() function can be used to extract the results.
scan <- lazyData(object = lg, dataset = "scan", pheno = pheno$pheno_names[1])
scan[which.max(scan$negLog10P), ]
```
  
<br>

The `callPeakBlock()` function defines peaks based on adjusted p-values (FDR) 
obtained from association analysis at markers, as well as genotype correlations 
between markers — in other words, linkage disequilibrium.
```{r message=FALSE}
callPeakBlock(object = lg, signif = 0.05, threshold = 0.8)
```
  
<br>

The peaks detected by the `callPeakBlock()` function can be visualized using
the `plotPeaks()` function.
```{r warning=FALSE}
for(i in seq_along(pheno$pheno_names)){
  p <- plotPeaks(object = lg, pheno = i)
  print(p)
}
```

<br><br>

## Recalculating and Visualizing Peaks
As shown in the peak plot above, visual inspection may reveal many small peak 
regions detected within a larger peak region, which might be considered a single 
peak. To group these small peaks into a single peak, you can use the 
`recalcAssoc()` function. This function tests the independence of peaks based on
multiple regression analysis. In the multiple regression model, all possible 
pairs of peaks are selected from the detected peaks, and each combination is 
tested to assess their independence by comparing the model fit of a full model
— which includes both selected peaks as variables — and a reduced model that 
includes only one of the peaks. If the full model is preferred over the reduced 
model at a p-value threshold (the default is 0.05), the two peaks are considered
independent. Otherwise, the higher peak (the one with the greater -log10(P)
value) is retained, and the other is removed and recorded as grouped with the 
higher peak in the `groups` dataset (companion store by default, or the GDS
`lazygas/recalc/groups/` subtree when `lazygas_store = "gds"`).
  
<br>

The refine_position argument specifies whether to recalculate the p-values of
markers around each peak using multiple regression models, treating peaks other
than the one under test as fixed-effect variables. This process may result in 
changes to the original peak positions defined by `callPeakBlock()`. 
The `grouping_threshold` argument specifies the threshold for grouping peaks. 
The default is 0.05.
```{r}
recalcAssoc(object = lg, 
            n_threads = 1, 
            refine_position = FALSE, 
            grouping_threshold = 0.05)
```
  
<br>

The recalculated peaks can be visualized using the `plotPeaks()` function with 
`recal = TRUE`.
```{r warning=FALSE}
for(i in seq_along(pheno$pheno_names)){
  p <- plotPeaks(object = lg, pheno = i, recalc = TRUE)
  print(p)
}
```

<br><br>

## Visualizing Haplotype and Phenotype Data
You can generate haplotype-wise phenotype distribution plots using the 
`haploPlot()` function for peaks both before and after applying `recalcAssoc()`.
Plot titles include peak location, `-log10P`, and when scan results are available,
the additive effect `Coef.add` with allele direction (original phenotype units
for continuous traits).
```{r}
for(i in seq_along(pheno$pheno_names)){
  out <- haploPlot(object = lg, pheno = i, recalc = FALSE)
  # As the output is a list of ggplot objects, only the first plot is shown here. 
  print(out[[1]]) 
}
```

<br>

```{r}
for(i in seq_along(pheno$pheno_names)){
  out <- haploPlot(object = lg, pheno = i, recalc = TRUE)
  # As the output is a list of ggplot objects, only the first plot is shown here. 
  print(out[[1]]) 
}
```

<br>
  
## List up candidate genes
The `listCandidate()` function helps you list candidate genes for the detected 
association peaks. You are required to provide gene annotation as a `GRanges` object (for example from `rtracklayer::import.gff()` or `rtracklayer::import.gff3()`); chromosome names must be consistent with the GDS.
If you have a VCF file output from SnpEff, you can also provide it 
to the function to summarize variant impacts on gene functions. Since a VCF file
output from SnpEff is large and takes a long time to load into the R 
environment, the `snpeff2gds()` function allows you to convert the VCF file into
a GDS file. The converted GDS file must then be provided to the `listCandidate()`
function after opening a connection to the file using the `open_snpeff()`
function. In addition, a data.frame containing other gene 
annotations can be provided to the `listCandidate()` function to include that
information in the output candidate list. The data.frame must contain a column 
named `Gene_ID`, which should match the gene IDs in the GFF (`gff$ID`).

### Example with your own annotation files
```{r eval=FALSE}
# Do not run.
gff_fn <- "path/to/gff/gene_annotation.gff"
gff <- rtracklayer::import.gff(gff_fn)

snpeff_fn <- "path/to/SnpEff/output/variants.snpeff.vcf"
gds_fn <- sub("\\.vcf", ".gds", snpeff_fn)
snpeff2gds(vcf_fn = snpeff_fn, out_fn = gds_fn)
snpeff_gds <- open_snpeff(gds_fn = gds_fn)

ann <- read.csv("path/to/gene_annotation.csv")
names(ann)[1] <- "Gene_ID"
listCandidate(object = lg, gff = gff, snpeff = snpeff_gds, ann = ann, recalc = TRUE)
```

### Example with bundled demo annotation
The package ships synthetic GFF, SnpEff VCF, and gene-description files that overlap
peaks detected in `sample.gds`. The following chunk lists demo candidate genes for
the trait analysed above.

```{r message=FALSE, warning=FALSE}
gff_fn <- system.file("extdata", "demo_annotation.gff", package = "lazyGas")
vcf_fn <- system.file("extdata", "demo_snpeff.vcf", package = "lazyGas")
ann_fn <- system.file("extdata", "demo_ann.csv", package = "lazyGas")

have_demo_ann <- nzchar(gff_fn) && file.exists(gff_fn) &&
  nzchar(vcf_fn) && file.exists(vcf_fn)

if (have_demo_ann) {
  gff <- rtracklayer::import.gff(gff_fn)
  ann <- read.csv(ann_fn, stringsAsFactors = FALSE)
  snpeff_gds_fn <- tempfile(pattern = "lazygas_snpeff_", fileext = ".gds")
  snpeff2gds(vcf_fn = vcf_fn, out_fn = snpeff_gds_fn, verbose = FALSE)
  snpeff_gds <- open_snpeff(snpeff_gds_fn)
  listCandidate(
    object = lg,
    gff = gff,
    snpeff = snpeff_gds,
    ann = ann,
    recalc = TRUE
  )
  candidate <- lazyData(object = lg, dataset = "candidate", pheno = "Fruit weight")
  knitr::kable(candidate[, c("peak_ID", "Gene_ID", "Gene_chr", "negLog10P", "Name")])
} else {
  message("Demo annotation files are not installed; skipping listCandidate() example.")
}
```

<br>

## Refining candidate genes (`searchCandidateGenes`)
After `listCandidate()`, the candidate table may still be large. The function `searchCandidateGenes()` filters and ranks rows using free-text annotation columns (for example descriptions or GO terms merged via the `ann` argument in `listCandidate()`).

- **`mode = "keyword"`** (only supported mode) — phrase-aware matching with English stopwords ignored and word-boundary / multi-word phrase patterns; with `keyword_match = "all"`, every content term must appear (default is `"all"`; use `"any"` for OR-style matching).

By default, all columns except peak/SnpEff coordinate columns are concatenated as annotation text; override with `ann_cols`. Use `object` and `pheno` to read candidates from the companion store, or pass a `data.frame` as `candidate`.

```{r}
cand_demo <- data.frame(
  peak_ID = 1L,
  Gene_ID = c("g1", "g2", "g3"),
  Gene_chr = "1",
  Gene_start = 1:3,
  dist2peak = 0,
  negLog10P = 3,
  Description = c(
    "fruit weight development",
    "root hair elongation",
    "cell wall biosynthesis"
  ),
  stringsAsFactors = FALSE
)
searchCandidateGenes(
  candidate = cand_demo,
  query = "fruit weight",
  mode = "keyword",
  keyword_match = "all"
)
```

```{r eval=FALSE}
searchCandidateGenes(object = lg, pheno = "Fruit weight",
                       query = "fruit ripening carbohydrate",
                       mode = "keyword", keyword_match = "any", top_n = 50)
```

<br><br>

## Extracting Data
The phenotype data assigned to a `LazyGas` class object, along with the output
data generated by functions in the `lazyGas` package, can be retrieved using the 
`lazyData()` function. The API is the same regardless of `lazygas_store`; only the
on-disk layout differs (Parquet folder, SQLite file, or GDS subtree).
```{r}
pheno <- getPheno(object = lg)
sapply(pheno, head)
```

<br>

The output of association analysis.
```{r}
scan <- lazyData(object = lg, dataset = "scan", pheno = pheno$pheno_names[1])
head(scan)
```

<br>

The output of peak calling.
```{r}
peakcall <- lazyData(object = lg, dataset = "peakcall", pheno = pheno$pheno_names[1])
head(peakcall)
```

<br>

The output of peak calling after peak grouping.
```{r}
recalc <- lazyData(object = lg, dataset = "recalc", pheno = pheno$pheno_names[1])
head(recalc)
```

<br>

The record of peak grouping.
```{r}
groups <- lazyData(object = lg, dataset = "groups", pheno = pheno$pheno_names[1])
head(groups)
```

<br>

The list of candidate genes.
```{r eval=have_demo_ann}
head(candidate)
```

<br>

The per-variant SnpEff table stored for the variant viewer.
```{r eval=have_demo_ann}
snpeff_tab <- lazyData(object = lg, dataset = "snpeff", pheno = "Fruit weight")
head(snpeff_tab[, c("Gene_ID", "Chr", "Pos", "Annotation", "Annotation_Impact")])
```

<br><br>

## Variant viewer
After `listCandidate()` with SnpEff annotation, you can inspect a single candidate
gene with `getVariantViewerData()`. The function merges stored SnpEff rows, scan
statistics, and GFF gene structure into a list that includes a ready-made
`ggplot` object.

```{r eval=have_demo_ann, fig.width=9, fig.height=4}
demo_gene <- candidate$Gene_ID[1]
vdata <- getVariantViewerData(
  object = lg,
  gene_id = demo_gene,
  gff = gff,
  pheno = "Fruit weight",
  snpeff = snpeff_gds,
  ann = ann,
  recalc = TRUE
)
plotVariantViewer(vdata)
```

The same plot can be produced in one step from a `LazyGas` object:

```{r eval=have_demo_ann, fig.width=9, fig.height=4}
plotVariantViewer(
  object = lg,
  gene_id = demo_gene,
  gff = gff,
  pheno = "Fruit weight",
  snpeff = snpeff_gds,
  ann = ann,
  recalc = TRUE
)
```

<br><br>
  
## Generating Interactive HTML Report
To organize the results of the analysis, you can generate an interactive HTML file.
`makeInteractiveSummary()` writes Manhattan plots, peak views, haplotype plots, and
candidate tables. `makeInteractiveDashboard()` adds variant-viewer panels below the
summary; clicking a `Gene_ID` in the candidate table scrolls to the corresponding
gene view.

The `what` argument accepts a character vector specifying one or more types of
plots to generate. Users can choose from the following options:
- `"scan"`: Draws an **interactive Manhattan plot**.  
- `"scan_png"`: Draws a **static Manhattan plot** (requires suggested package **base64enc**).  
- `"peakcall"`: Draws an **interactive peak call plot**.  
- `"recalc"`: Draws an **interactive recalculated peak call plot**.  
- `"groups"`: Draws an **interactive peak grouping list**.  
- `"peakcall_haplo"`: Draws an **interactive haplotype-wise phenotype distribution plot** for peaks **before recalculation**.  
- `"recalc_haplo"`: Draws an **interactive haplotype-wise phenotype distribution plot** for peaks **after recalculation**.  
- `"candidate"`: Draws an **interactive list of candidate genes**.

### Summary report
```{r eval=FALSE}
makeInteractiveSummary(
  object = lg,
  pheno = "Fruit weight",
  out_fn = "sample_summary.html",
  what = c("scan", "scan_png", "peakcall", "recalc", "groups", "candidate")
)
```

### Dashboard with variant viewer
Run `listCandidate()` first (as in the demo section above), then:

```{r eval=FALSE}
makeInteractiveDashboard(
  object = lg,
  pheno = "Fruit weight",
  gff = gff,
  out_fn = "lazygas_dashboard.html",
  snpeff = snpeff_gds,
  ann = ann,
  recalc = TRUE,
  what = c("scan_png", "peakcall", "recalc", "candidate")
)
```

Open `lazygas_dashboard.html` in a browser. The `Gene_ID` column contains links;
click a gene ID to reveal its variant viewer panel.

### One-command demo
To rebuild the sample analysis and write a dashboard without assembling the
steps manually:

```{r eval=FALSE}
runDashboardDemo(out_dir = "demo_output")
# Opens: demo_output/lazygas_dashboard.html
```

From the shell, after installing or loading the package from source:

```{bash, eval=FALSE}
Rscript inst/demo/run_dashboard_demo.R demo_output
```

### MVP feature demo (v0.6+)

`runMvpDemo()` exercises the newer APIs on bundled data: three traits
(`demo_pheno_multitrait.csv`), gene annotation, and SnpEff effects. It runs
`runLazyGas()`, prints GWAS QC and cross-trait peak summaries, conditional
association and credible-set results, then writes an HTML report.

```{r eval=FALSE}
runMvpDemo(out_dir = "demo_output/mvp")
# -> demo_output/mvp/lazygas_mvp_report.html
# Companion store: demo_output/mvp/sample.lazygas/
```

From the shell:

```{bash, eval=FALSE}
Rscript inst/demo/run_mvp_demo.R demo_output/mvp
```

Bundled MVP inputs in `inst/extdata/`:

| File | Role |
|------|------|
| `sample.gds` | Genotypes (100 samples, chromosomes 1–12) |
| `demo_pheno_multitrait.csv` | `fruit_weight`, `fruit_length`, `flowering_time` |
| `demo_annotation.gff` | Gene models for candidate listing |
| `demo_snpeff.vcf` | SnpEff annotations (converted to GDS on first run) |
| `demo_ann.csv` | Gene names and descriptions |
| `demo_orthologs.csv` | Ortholog IDs/scores for evidence ranking (`collectGeneEvidence()`, `rankPhenotypeCandidates()`) |
| `demo_expression.csv`, `demo_expression_meta.csv` | Expression evidence for `runExplorerDemo()` |
| `pheno.csv` | Single-trait demo phenotype for this vignette |

### MVP workflow: runLazyGas, QC, and cross-trait view

Run the core pipeline in one call after assigning phenotypes:

```{r eval=FALSE}
conv_fun <- makeConvFun(geno_format = "dosage", n_levels = 3)
lg <- runLazyGas(
  object = lg,
  steps = c("scan", "peakcall", "recalc", "candidate", "dashboard"),
  gff = gff,
  snpeff = snpeff_gds,
  ann = ann,
  out_fn = "lazygas_dashboard.html",
  formula = "add + dom",
  conv_fun = conv_fun,
  geno_format = "dosage",
  resume = TRUE
)
```

GWAS quality control:

```{r eval=FALSE}
summarizeGWASQC(lg, pheno = "Fruit weight")
plotQQ(lg, pheno = "Fruit weight")
```

When multiple traits are assigned, compare shared peaks by genomic distance and
peak-marker genotype correlation:

```{r eval=FALSE}
plotMultiTraitOverview(lg, recalc = TRUE)
clusterCrossTraitPeaks(lg, dist_threshold = 500000, r2_threshold = 0.6)
```

Fine-mapping within a peak (conditional scan and credible set):

```{r eval=FALSE}
# Run per-peak conditional + credible-set analysis for all peaks
runFineMapping(lg, pheno = "Fruit weight", recalc = TRUE)

# Interpretation-focused summary (stored in sample.lazygas/fine_mapping/)
summarizeFineMapping(lg, pheno = "Fruit weight", recalc = TRUE)
```

Per-peak detail:

```{r eval=FALSE}
conditionalAssoc(lg, pheno = "Fruit weight", peak_id = 1, recalc = TRUE)
summarizeConditionalSignals(lg, pheno = "Fruit weight", peak_id = 1)
calcCredibleSet(lg, pheno = "Fruit weight", peak_id = 1, use_conditional = TRUE)
plotCredibleSet(lg, pheno = "Fruit weight", peak_id = 1)
```

Include fine-mapping tables in HTML reports with `what = "fine_mapping"` in
`makeInteractiveSummary()` or `runLazyGas(..., steps = c(..., "fine_mapping", "summary"))`.

### Interpreting fine-mapping results (important)

Phase-2 fine-mapping in lazyGas is designed for **interpretation support**, not
for claiming that a specific SNP is the causal variant.

**What conditional analysis tells you**

- `summarizeConditionalSignals()` counts **independent association signals**
  within a peak (stepwise conditioning).
- `n_independent_signals = 1` (or 0 below threshold): one association blob in the
  LD block — typical for small mapping populations.
- `n_independent_signals >= 2`: consider separate follow-up per lead variant
  (different haplotype backgrounds or multiple QTLs in the same region).

Conditional analysis does **not** identify the causal SNP; it separates correlated
tags from potentially independent signals.

**What credible sets tell you**

- `calcCredibleSet()` uses Wakefield ABF under a **single-causal-variant**
  approximation and does **not** model LD explicitly.
- `resolution` in `summarizeFineMapping()`:
  - `high` (≤3 variants in the 95% set): locus narrowed within the LD block
  - `moderate` (4–10): partial narrowing
  - `low` (>10): **not finely resolved** — treat as uncertainty, not SNP identity
- A **large credible set** usually means limited sample size, strong LD, and/or
  weak per-SNP separation — the same information you would already infer from a
  broad GWAS peak.
- **Top PIP ≈ marginal GWAS lead** is expected with ABF; a different top-PIP SNP
  is rare and still lies in the same LD neighbourhood.

**Practical guidance (especially N ≈ 200–300 in non-human studies)**

1. Use fine-mapping to **flag unresolved loci** (`resolution = "low"`) before
   investing in variant-level functional tests.
2. Prefer **gene-level follow-up** (`listCandidate()`) when credible sets span
   many SNPs.
3. Do not equate PIP or “95% credible set” with a 95% chance that the causal
   variant is identified ([Hutchinson & Wallace 2020](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1007829)).
4. For causal-variant claims, plan **independent populations**, higher recombination
   (MAGIC/NAM), or functional validation — not finer PIP thresholds alone.

**What lazyGas fine-mapping is not**

- Not a replacement for SuSiE/FINEMAP (no explicit LD-aware multi-SNP model).
- Not colocalization (`coloc`); trait-side PIP for eQTL integration is out of scope here.
- Not proof of causality — a reporting aid for peak blocks you already detected.

### Phenotype explorer and local LLM (Apptainer, official setup)

Since v0.7.0, lazyGas can rank candidate genes by phenotype-guided evidence and
optionally use a **local LLM** (Ollama) for query parsing and natural-language
explanations. Ranking and evidence collection are deterministic in R; the LLM is
optional. **v0.8.0** documents GWAS effect sizes in input trait units.
**v0.9.0** adds Shiny-embedded GWAS overview and on-demand locus / variant
views (alongside the existing HTML `makeInteractiveDashboard()` reports).
**v0.9.1** adds a thin **pipeline runner** Shiny app so scan through candidate
listing can be launched from the GUI without mixing those steps into the
explorer UI. **v0.9.2** further reorganizes explorer controls by tab and adds a
dedicated **Local LLM** tab with runtime/model diagnostics.

**Official environment:** lazyGas ships an **Apptainer** workflow so you do not
need a host `ollama` binary (typical on HPC clusters). Models and cache are stored
on the host under `lazyGasOllamaHome()` (default:
`tools::R_user_dir("lazyGas", "cache")/ollama`, or set `LAZYGAS_OLLAMA_HOME`).

**Prerequisites**

- [Apptainer](https://apptainer.org/) or Singularity on `PATH`
- Network access for the first `docker://ollama/ollama` image pull
- Suggested: **processx** (background `ollama serve` from R)

**One-time setup from R**

```{r eval=FALSE}
library(lazyGas)

# Pull Ollama image and install the lazygas-ollama CLI wrapper
configureLazyGasOllama(model = "llama3.2:3b")

# Start server and download the model (first run may take several minutes)
startLocalLLM(model = "llama3.2:3b")
```

`configureLazyGasOllama()` creates:

| Artifact | Default location |
|----------|------------------|
| Apptainer image (`.sif`) | `{lazyGasOllamaHome()}/ollama.sif` |
| Model cache (bind mount) | `{lazyGasOllamaHome()}/` → `/root/.ollama` in container |
| CLI wrapper | `{lazyGasOllamaHome()}/bin/lazygas-ollama` |

Environment variables set for the session: `LAZYGAS_OLLAMA_SIF`, `LAZYGAS_OLLAMA_HOME`,
`LAZYGAS_OLLAMA_CLI`, `LAZYGAS_LLM_URL`, `LAZYGAS_LLM_MODEL`.

**Shell alternative** (same image as R):

```bash
bash inst/apptainer/build_ollama_sif.sh
export LAZYGAS_OLLAMA_SIF="$HOME/.cache/lazygas/ollama/ollama.sif"
```

Then in R: `configureLazyGasOllama(build_sif_if_missing = FALSE)`.

**Two-app workflow (v0.9.1+)**

Use **`runLazyGasRunner()`** once to build companion-store results from GDS,
phenotype CSV, and GFF paths on disk, then **`runLazyGasExplorer()`** to
explore and rank candidates. `runExplorerDemo()` runs the full R pipeline and
is equivalent to using the runner plus explorer on bundled demo files.

```{r eval=FALSE}
# Option A: CLI demo (writes demo_output/explorer/)
runExplorerDemo(out_dir = "demo_output/explorer", overwrite = TRUE)
runLazyGasExplorer()   # GDS path: demo_output/explorer/sample.gds

# Option B: GUI pipeline, then explorer
runLazyGasRunner()     # set GDS / pheno CSV / GFF paths, Run pipeline
runLazyGasExplorer(gds_path = "demo_output/explorer/sample.gds")
```

In the Shiny UI, open the **Local LLM** tab to run **Setup Ollama
(Apptainer)** once, then **Start Ollama** / **Pull model**. Chat and
comparisons work without the LLM (rule-based fallback); enable
**Use local LLM** for richer parsing and explanations.

After **Load project**, the **GWAS overview** tab is the primary place to set
the active phenotype and review phenotype/Manhattan/peak plots (static PNG
Manhattan by default). Use **Ranked genes** for phenotype-query inputs and
ranking. After ranking, select a gene and open **Locus / variants** for
haplotype and on-demand variant-viewer plots (provide a GFF path; demo default
is `inst/extdata/demo_annotation.gff`).

During long-running **Rank candidates**, **Generate explanation**, and **Ask**
actions, the explorer shows progress and temporarily disables action buttons to
avoid duplicate submissions.

**Diagnostics**

```{r eval=FALSE}
describeOllamaStatus()
listOllamaModels()
stopOllamaServer()   # stops only processes started from this R session
```

Host-installed Ollama (https://ollama.com) still works when `ollama` is on `PATH`
and no Apptainer wrapper is configured.

The Shiny apps and Quarto templates ship under `inst/shiny/runner/`,
`inst/shiny/explorer/`, and `inst/quarto/`. Launch the pipeline runner with
`runLazyGasRunner()` and the phenotype explorer with `runLazyGasExplorer()`.

<br><br>

```{r}
sessionInfo()
```

